A Logical Solution of the Linearity Challenge

### 2025-08-02 paper/introduction.tex
++

This paper reports on a solution to one of the challenges in the benchmark --
hereafter called \emph{linearity challenge} -- which is concerned with the
formalization of a minimal \emph{session calculus}. Sessions and session
types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16} are established
formalisms for the static analysis of distributed programs based on peer-to-peer
communication channels. Every session type system revolves around three key
ideas: (1) session endpoints are \emph{linear resources} that cannot be
discarded or duplicated without jeopardizing some key safety and liveness
properties of a program; (2) the type of a session endpoint \emph{must be
updated} after each use to reflect the state of the interaction it describes;
(3) peer session endpoints are meant to be used in complementary ways so as to
guarantee the absence of communication errors and, to some extent, progress of
the interaction.

The calculus that we formalize -- called \Calculus in the following -- and the
one in the description of the challenge differ for two main aspects:
\begin{enumerate}
\item The challenge considers a minimal calculus of sessions, which is basically
    a first-order $\pi$-calculus without replication, whereas \Calculus is
    closely related to
    \CP~\citep{CairesPfenning10,Wadler14,CairesPfenningToninho16}, a
    fully-featured calculus of higher-order sessions with both linear and shared
    channels and whose type system is based on linear logic.
    %
    This allows us to target a model of sessions that is well-known in the
    literature, that is based on firm logical grounds, and that includes a
    broader range of features.
\item While the calculus in the challenge is equipped with native sessions,
    \Calculus is a \emph{linear} calculus where each channel is meant to be used
    for a \emph{single} communication. Structured conversations pertaining the
    same session and involving the exchange of several messages can still be
    modeled thanks to the use of \emph{explicit continuation channels}, in the
    same spirit of the encoding of (binary) sessions into the linear
    $\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
    %
    Overall, \Calculus has the same expressive power of \CP but is a lower-level
    calculus in which sessions can be encoded instead of being featured
    natively.
\end{enumerate}

We value these differences because their combined effect allows us to obtain a
remarkably simple formalization of \Calculus.
%
The fact of working with a calculus based on linear logic prevents \emph{by
construction} the same (sequential) process to own both endpoints of a session.
This situation is clearly undesirable since it does not correspond to a useful
pattern of interaction (every meaningful session requires its endpoints to be
used by parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of polarities~\citep{GayHole05} or
by using different names connected by the same
binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Also, the fact of working with a calculus with \emph{linear} channels and
explicit continuations spares us the need to \emph{update} the type of channels
in typing contexts. When a channel is used it is effectively consumed, therefore
its type is \emph{removed} from the typing context and the type of the
continuation channel is \emph{added} back to the typing context. As is turns
out, removing and adding types is easier than updating them. Even more so
considering that the type of continuation channels can be conveniently added
\emph{at the beginning} of a typing context, since such channels are fresh by
definition.

>>>>

This paper reports on a solution to one of the challenges in the benchmark
concerned with the formalization of a minimal \emph{session calculus}. Hereafter
we will refer to this as the \emph{linearity challenge}.
%
Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
are established formalisms for the static analysis of distributed programs based
on peer-to-peer communication channels. Every session type system revolves
around three key ideas: (1) session endpoints are \emph{linear resources} that
cannot be discarded or duplicated without jeopardizing some safety and liveness
properties of a program; (2) the type of a session endpoint is \emph{updated}
after each use to reflect the state of the interaction it describes; (3) peer
session endpoints are meant to be used in complementary ways so as to guarantee
the absence of communication errors and, to some extent, progress of the
interaction.

The calculus that we formalize, which we call \Calculus, and the one in the
description of the challenge differ for two main aspects:
\begin{enumerate}
\item The challenge considers a minimal calculus of sessions, which is basically
    a first-order $\pi$-calculus without replication, whereas \Calculus is
    closely related to
    \CP~\citep{CairesPfenning10,Wadler14,CairesPfenningToninho16}, a
    fully-featured calculus of higher-order sessions with both linear and shared
    channels and whose type system is based on linear logic.
    %
    Targeting \Calculus allows us to consider a model of sessions that is
    well-known in the literature, that is based on firm logical grounds, and
    that includes a broader range of features including higher-order sessions
    and shared channels.
\item While the calculus in the challenge is equipped with native sessions,
    \Calculus is a \emph{linear} calculus where each channel is meant to be used
    for a \emph{single} communication. Structured conversations pertaining the
    same session and involving the exchange of several messages can still be
    modeled thanks to the use of \emph{explicit continuation channels}, as in
    the encoding of (binary) sessions into the linear
    $\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
    %
    Overall, \Calculus has the same expressive power of \CP but is a lower-level
    calculus in which sessions can be encoded instead of being featured
    natively.
\end{enumerate}

\Luca{Da qualche parte dire quali sono le sfide della sfida: representation of
channel names, context splitting. Parlare di co de bruijn?}

We value these aspects of \Calculus because their combined effect allows us to
obtain a remarkably simple formalization.
%
The fact of working with a calculus based on linear logic prevents \emph{by
construction} the same (sequential) process to own both endpoints of a session.
This situation is clearly undesirable since it does not correspond to a useful
pattern of interaction (every meaningful session requires its endpoints to be
used by parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of polarities~\citep{GayHole05} or
by using different names connected by the same
binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Also, the fact of working with a calculus with \emph{linear} channels and
explicit continuations spares us the need to \emph{update} the type of channels
in typing contexts. When a channel is used it is effectively consumed, therefore
its type can be \emph{removed} from the typing context and the type of the
continuation channel is \emph{added} back to the typing context. As is turns
out, removing and adding types is easier than updating them. Even more so
considering that the type of continuation channels can be conveniently added
\emph{at the beginning} of a typing context, since such channels are fresh by
definition.
